Limited predictability of amino acid substitutions in seasonal influenza viruses

### 2020-12-18 response_to_referees.tex

\refa{1. The prediction problem is defined for each mutation: based on the
tracked frequency trajectory, can the future (fixation, loss, or
polymorphism) of the mutation be predicted. Such formulation has been
previously proposed by Illingworth and Mustonen, (eg. Genetics 2011,
Plos Pathogens 2012). By averaging over all amino-acid substitutions,
the authors show that the fates of mutations are not determined by the
value of the starting frequency.}

\response{We thank the reviewer for pointing out the work by Illingworth and Mustonen, which is indeed very relevant and complementary to our approach.
We discuss their relation below. }

\refa{1.1. Mutations in influenza are highly nested, with a substantial
hitch-hiking, and no attempt is made to disentangle such dependencies when counting the mutations.}

\response{
The reviewer is correct that our original manuscript did not attempt to account for nested mutations and we agree that this should have been done. \\
We added a new section in the Supplementary Material (``Correcting for nested trajectories'') where investigate the extent to which nesting affects our results.
We cluster trajectories of mutations that partly appear on the same strains to various degrees, and trajectories corresponding to mutations always or often appearing on the same strains are then counted as one ``effective" trajectory.
Our analysis shows that nesting of mutations does not change our results significantly.
We therefore left the figures of main text unchanged and refer to the supplement for a discussion of this issue.
}

@@

\refa{1.2 Despite the more general formulation in the beginning, this
approach makes use only of the last time point in the trajectory,
rather than the full trace (contrary to the work of Illingworth\&
Mustonen). I find the observed neutral-like statistics not surprising
for such a limited data input, which doesn't capture past frequency dynamics. Therefore, these conclusions should be
revisited and benchmarked against the more general implementation of Illingworth\&Mustonen-like approach.}

\response{
	Illingworth and Mustonen (I\&M) presented a careful analysis to what extent trajectories can be fit using a model assuming a single fitness coefficient, or a model that differentiates different background fitnesses.
	Our results are consistent with their finding that a single parameter model (which only allows monotonic trajectories) does not capture the majority of trajectories.
	Beyond that, our approach and aim differ from I\&M.
	We investigate \emph{a priori} features of trajectories that potentially have information on whether the mutation fixes, not whether the trajectories can be parametrized by a model.
	Fitting a model to a partial trajectory to predict fitness is not expected to work: I\&M showed that subsequent interference has a substantial effect and our results indicate that the speed at which the mutation rises does not predict fitness.

	Furthermore, I\&M estimate fixing mutations to have a much higher intrinsic fitness than dying mutations while we show that very little such information is available a priori.
	The retrospective fit to trajectories tends to assign a positive fitness effect to fixing (and hence on average rising) mutations.
	This is why we resort to a model free approach to ask what a priori information can help to determine which mutations fix or die.
\responsedraft{Should we mention all the changes we made in the text? 
We spend more time explaining that even though there is linkage, and even though we're looking at just the last point, we expect something above diagonal. 
(Last paragraph before "Results", last paragraph before "Predicting future frequencies"), and first paragraph of "Simulations of models of adaptation"}

@@

% 	The second reason is that we believe that our model-free approach has intrinsic benefits. [Something saying that we'd prefer not to start fitting models to  data, because it's not the point. But I'm not sure how to formulate this.] \\
% 	Finally, we are aware that using only the last time point of a rising trajectory gives only limited opportunity for predicting its future. However, observing apparently-neutral statistics in the case of A/H3N2 is still surprising to us. The case of simulated populations shows that clear signs of selection can be observed by looking at basic features of trajectories (\emph{e.g.} figure S20 of the supplementary material). This is consistent with the theory of evolving population: if the only available information about a mutation is that it has risen in frequency starting from $0$, it has a higher chance on being beneficial than deleterious, and its fixation probability should be higher than what it would be in a neutral scenario. Hence, even with our very simple approach, we observe that predictability of A/H3N2 qualitatively differs from what models of adapting populations would suggest. \\}

\refa{2. The authors examined the predictive power of one predictive
method, the LBI. However, they did not do a systematic comparison of
the different methods that they cite (Morris et al, 2018), which differ in prediction
targets and methods. Therefore, the general conclusions about
predictability, eg. on page 8, in line 47, and on page 9, line 48 are
too sweeping and should be made precisely for those methods looked at
in detail.}

\response{The reviewer is right to point out that our original manuscript did not perform a systematic comparison of different prediction methods.

In a recent model-driven effort (Huddleston et al), we investigate the power of sequence based, antigenically informed, experimentally informed, or phylogenetic (LBI) scores to predict future influenza virus populations.
The LBI was among the highest performing models and we hence used this score in the original manuscript.
But we took the suggestion of the referee (and referee 2) to heart and now include all highly performing models from Huddleston \emph{et al.} in a new section in the Supplementary Material ("Ability of fitness models to predict fixation"). 
These additional results show that only mutational load has some power to differentiate between fixing and non-fixing trajectories.
We believe this addition has made the manuscript stronger and we thank both referees for the suggestion.

We also agree with the reviewer that these different methods differ in prediction targets: our results indicate that most of them are not predictive of frequency dynamics, but they remain predictive of the composition of the future population. 
In consequence, we have toned down our conclusions about predictability. \pierre{I say that because I changed p8 l47. I have not touched p9 l48 because the new appendix shows it is a correct statement.}
}

\refa{3. Going through the previous and cited literature, I think the authors should cite some of these works in a more careful way. Specifically, the very related work by Illingworth and Mustonen is not mentioned. The distribution $P_delta_t(f|f_0)$ has been at the core of
the method of Strelkowa and Lassig, 2012 (which is cited, but at other
parts of the text), and the same paper also uses a similar simulation model, which should be acknowledged at the appropriate points of the text.}
\responsedraft{I cited (Strelkowa and Lassig) in the first paragraph of the section "Predicting future frequencies". 
Did not cite it close to the simulation part, will do. 
We cite Illingworth and Mustonen plenty of times now. 
}

@@

\refb{The paper does a retrospective study of amino acid substitutions in seasonal Influenza to determine what properties of these substitutions could help predict their fate in the future. The authors find that future frequency trajectories are surprisingly unpredictable. Even predicting which mutations fix in the population is hard. The authors find that the current frequency of a mutation is the best predictor for the probability of fixation, which would be expected under neutrality but not in a model with selection.
I appreciate this study, I think it will be of interest to many readers and it is generally well done and fairly easy to follow.
}

\response{We thank the reviewer for this assessment and the constructive suggestions below. }

\refb{1. The authors focus on one feature at a time (frequency, epitope status, LBI …). It is interesting to see that each of these is not very predictive of future frequency / prob of fixation. However, I think the obvious next step is to see whether a combination of many features could do a better job of predicting. I am not sure why the authors don’t try to fit a model that takes into account all information they have about sites (say, type of AA change, location in the gene, current frequency etc) and see if a ML model is able to make predictions. }

\response{
	A similar point was also brought up by reviewer 1.
	In addition to the LBI, epitopes, age, and geographic distribution we now also include the fitness scores of the best performing models in a recent machine learning effort to predict the composition of future populations (as opposed to predicting fixation).
	These models include antigenic, experimental, and sequence or phylogenetic information.
	Interestingly, all these models do similarly poorly with the exception of mutational load.
	These results are now included as supplementary information and discussed.
}

@@

\refb{2. Fig 2A and 2B look quite different to me. In the text it appears to me as if these are very similar to the authors. What are the characteristics of the AAs that fix in H1N1? }

\response{Whether 2A or 2B are perceived as similar or not probably depends on how big a deviation from the diagonal one expected a priori.
We now discuss now highlight the differences in the text and discuss potential reasons for the different behaviour.  }

@@

\refb{3. Question: what do these results mean for vaccines? How to decide which strains to use for vaccines? This may be known to those who work on Influenza, but for a relative outsider it is not clear. }

\response{
	Computational prediction of which circulating strains will dominate future influenza populations have been attempted for about 2 decades and have recently become a regular input to the biannual consultations on the influenza vaccine composition.
	Since performance of predictive models can only be assessed on limited historical data, we believe that our model-free approach clarifies which aspect of influenza evolution are predictable and which ones are not.
}
\responsedraft{We should maybe make clear that this is not directly related to predictions for vaccines (at least it's not the aim). 
That's John's paper. 
The main implications are about our understanding of the evolutionary dynamics of influenza. 
This could also help us to reply to the first comment: doing ML is more useful to predict the future population (for vaccines) than for predicting fixation of a single mutation. }

### 2020-12-18 response_to_referees.tex

% 	Finally, we are aware that using only the last time point of a rising trajectory gives only limited opportunity for predicting its future. However, observing apparently-neutral statistics in the case of A/H3N2 is still surprising to us. The case of simulated populations shows that clear signs of selection can be observed by looking at basic features of trajectories (\emph{e.g.} figure S20 of the supplementary material). This is consistent with the theory of evolving population: if the only available information about a mutation is that it has risen in frequency starting from $0$, it has a higher chance on being beneficial than deleterious, and its fixation probability should be higher than what it would be in a neutral scenario. Hence, even with our very simple approach, we observe that predictability of A/H3N2 qualitatively differs from what models of adapting populations would suggest. \\}

\refa{2. The authors examined the predictive power of one predictive
method, the LBI. However, they did not do a systematic comparison of
the different methods that they cite (Morris et al, 2018), which differ in prediction
targets and methods. Therefore, the general conclusions about
predictability, eg. on page 8, in line 47, and on page 9, line 48 are
too sweeping and should be made precisely for those methods looked at
in detail.}

\response{The reviewer is right to point out that our original manuscript did not perform a systematic comparison of different prediction methods.

In a recent model-driven effort (Huddleston et al), we investigate the power of sequence based, antigenically informed, experimentally informed, or phylogenetic (LBI) scores to predict future influenza virus populations.
The LBI was among the highest performing models and we hence used this score in the original manuscript.
But we took the suggestion of the referee (and referee 2) to heart and now include all highly performing models from Huddleston et al.
These additional results show that only mutational load has some power to differentiate between fixing and non-fixing trajectories.
We believe this addition has made the manuscript stronger and we thank both referees for the suggestion.

### 2020-12-21 article.tex
small edits

Since genomic segments of influenza do not recombine, concurrent trajectories of mutations on the same segment will be correlated.
%Indeed, the trajectories corresponding to two mutations that partly appear on the same background will tend to show a similar behavior.
In the extreme case were the two mutations always appear on the same strains, their two trajectories will be identical.
This ``nested'' characted or frequency trajectories may introduce biases when performing statistical analysis.
In section \ref{sub:correcting_for_nested_trajectories} of the SM, we investigate the effect of such nested trajectories by clustering them based on the strains that compose them.
While the number of trajectories that contribute to the analysis decreases with more aggressive clustering, it does not significantly change the results presented below.

We focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$.
These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
Such distinction in novel and ancestral variants is necessary to meaningfully interrogate predictability.
Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one.
%If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants.
In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fix.
The sweep of a mutation can be due to its own fitness effect, to the genetic background or to the effect of the seven other segments.
By considering the ensemble of novel variants that are rising in frequency, we effectively average over backgrounds, obtaining a set of mutations that we expect to be beneficial on average.
If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.
Interference between multiple such advantages mutations will perturb the trajectories \citep{illingworth_components_2012} but we still expect an average \pierre{replaced ``consistent'' by ``average''} tendency to rise in frequency (see below). \pierre{Is this ``see below'' necessary? I think that's what the reader naturally does! }

>>>>

Since genomic segments of influenza do not recombine, concurrent trajectories of mutations on the same segment will be correlated.
%Indeed, the trajectories corresponding to two mutations that partly appear on the same background will tend to show a similar behavior.
In the extreme case were the two mutations always appear on the same strains, their two trajectories will be identical.
This ``nesting'' of frequency trajectories may introduce biases when performing statistical analysis.
In section \ref{sub:correcting_for_nested_trajectories} of the SM, we investigate the effect of such nested trajectories by clustering them based on the strains that compose them.
While the number of trajectories that contribute to the analysis decreases with more aggressive clustering, it does not significantly change the results presented below.

We focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$.
These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
Such distinction in novel and ancestral variants is necessary to meaningfully interrogate predictability.
Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one.
%If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants.
In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., they rise rapidly and fix.
The sweep of a mutation can be due to its own fitness effect, to the genetic background or to the effect of the seven other segments.
By considering the ensemble of novel variants that are rising in frequency, we effectively average over backgrounds, obtaining a set of mutations that we expect to be beneficial on average.
If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.
Interference between multiple such advantages mutations will perturb the trajectories \citep{illingworth_components_2012} but we still expect that they rise in frequency on average.
